Northern Pikeminnow Predation on Juvenile Chinook Salmon and Potential Impacts to Local Populations in the Upper Salmon River, Idaho

### 2021-12-20 analysis/paper/paper.Rmd
bioenergetic methods edits

To estimate the total consumption potential (i.e., number/amount of juvenile consumed during the peaks of DSR [fall] and NRR [spring] emigrations) of Northern Pikeminnow in Deadwater Slough, we used the Fish Bioenergetics v4.0 application developed by @Deslauriers2017 applied in the R statistical software [@RCoreTeam2021]. The daily rate of consumption in grams for an individual Northern Pikeminnow was estimated based on predator and prey energy densities, predator start and end weights, and water temperatures. Predator energy density for Northern Pikeminnow was fixed at 6,703 Joules(J)/g [@Deslauriers2017]. Prey energy densities were fixed at 21,500 J/g for juvenile Chinook Salmon based on an estimate from @Moss2016 and 3,000 J/g for invertebrates. Additional settings included 0% assumed indigestible prety, and a variation in the proportion of diet that was invertebrates or fish ranging from 10% to 90% for both categories. We used a weight-length formula [@Parker1995] to calculate the average predator start weight from the average length of Northern Pikeminnow captured during our study using the FSA package in R [@Ogle2021]. The average length of Northern Pikeminnow captured at the Deadwater Slough during the spring and fall were 352.9 and 393.7 mm, respectively, which calculates to average starting weights of 504.4 and 670.5 g. Continuous water temperature data from 2019 was used in two alternative models, one using the baseline start and end weights and the second with a 10% increase in average end weight on Northern Pikeminnow.

We ran three separate bioenergetic models for this study. For the first model, we used temperature data from a 78-day period spanning September 1 through November 17 when DSR emigrants are known to enter the mainstem Salmon River from natal tributaries (e.g., Lemhi River) and begin their downstream migration.<!-- Did we just use data from Lemhi here? All tributaries? Was this subjective? --> During this time, water temperatures typically exceed the range (0-7$^\circ$C) that would illicit concealment behavior or torpor from juvenile Chinook Salmon<!--citation?-->. This first model assumed no growth for Northern Pikeminnow. The second model was instead run for one full year, but also assumed no growth for Northern Pikeminnow. The assumption of zero growth was included to estimate consumption if individual Northern Pikeminnow were not growing and/or the population is stable. Finally, the third model shows how an increase of 10% body weight or an increase in population size changes the estimates of consumption over a full year period.

To estimate the total biomass consumed, we then multiplied the biomass consumed by an individual Northern Pikeminnow for the given time period by the estimated population size for that time period. The total biomass was further converted to an estimate of the total number of Chinook Salmon consumed by dividing by a weight of 12 g per individual. This weight was an average from juvenile Chinook Salmon captured at six rotary screw traps throughout the operation season March to November in the Upper Salmon above Deadwater Slough including traps in the Lemhi, Pahsimeroi, and North Fork Salmon rivers and one trap operating near the Sawtooth hatchery.

<!--
A few things that need to be clarified in this section:
* Why were the dates Sept 1 - Nov 17 chosen? They make sense, but we should include our rationale
* Where did the temperature data come from? Hobo? Stream gauge? The methods say temperature data from 2019 was used, but the results seem to use 2020? 
* Why does it look like we only ran it for dates in 2020? Is that the only year we had temperature data for? Would the results look different if we used temperature data from a different year?
* The lengths of pikeminnow: did we only use pikeminnow we caught in the fall to get average length? Is there a difference in lengths between 2019 and 2020? 
* There needs to be more explanation of the inverts vs. fish. What proportion went to each and why? Did it change through time? Based on literature at all? 
* Where is the data that was used to get a weight of 12 g for Chinook emigrant? Was that for Chinook presmolts, or a combination of smolts and presmolts? If the latter, can we filter it to look at the average weight for presmolts only? And maybe smolts only for a different analysis? 
* Could we run a similar model for the spring emigration period? Figure out the dates, look at lengths of pikeminnow from the spring, convert to average weight, use average size of Chinook smolts (not presmolts) from all the Upper Salmon traps? Kevin can work on converting the spring catch into some kind of estimate of pikeminnow at Deadwater in the spring
-->

>>>>

To estimate the total consumption potential (i.e., the number or amount of juvenile Chinook Salmon consumed during the peaks of DSR [fall] and NRR [spring] emigrations) of Northern Pikeminnow in Deadwater Slough, we used the Fish Bioenergetics v4.0 application developed by @Deslauriers2017 and applied in the R statistical software [@RCoreTeam2021]. The daily rate of consumption in grams for an individual Northern Pikeminnow was estimated based on the following inputs: predator and prey energy densities, predator start and end weights, and water temperatures. Predator energy density for Northern Pikeminnow was fixed at 6,703 Joules(J)/g [@Deslauriers2017]. Prey energy densities were fixed at 21,500 J/g for juvenile Chinook Salmon [@Moss2016]. We used a weight-length formula [@Parker1995] to calculate the predator start weight using the average length of Northern Pikeminnow captured in Deadwater Slough during our study. The average length of Northern Pikeminnow during the fall and spring were 352.9 and 393.7 mm, respectively, which calculated to average starting weights of 504.4 and 670.5 g. The assumption of zero growth was included to estimate consumption if individual Northern Pikeminnow were not growing and/or the population is stable, and so start and end weights are equal. Mean daily water temperatures were summarized from 15-minute interval temperature readings available March 3, 2013 to June 14, 2021 from a gage station approximately 22 river kilometers downstream of Deadwater Slough near Shoup, Idaho [@USGS2016].

Our largest uncertainty for model parameters was the proportion of the Northern Pikeminnow diet consisting of fishes (e.g., juvenile Chinook Salmon) versus other food items like invertebrates. Therefore, we varied the proportion of the diet that was fish or invertebrates ranging from 10-90%, in 10% intervals, for both categories. Energy densities for invertebrates was fixed at 3,000 J/g. Separate models were also run for a 77-day fall period from September 15 - November 30 and a 92-day period in the spring from March 1 - May 31 to coincide with peak emigrations of DSR and NRR juveniles from the Lemhi River, the largest Chinook Salmon population in the Upper Salmon MPG. This resulted in 18 separate model runs (2 seasons and 9 diet scenarios). Each model run provided an estimate of the amount (grams) of fish consumed by an individual Northern Pikeminnow during the fall or spring time periods.

<!-- During this time, water temperatures typically exceed the range (0-7$^\circ$C) that would illicit concealment behavior or torpor from juvenile Chinook Salmon. -->

To estimate the total biomass of fish potentially consumed by Northern Pikeminnow in Deadwater Slough, we then multiplied the biomass consumed by an individual Northern Pikeminnow for the time period by the estimated population size for that time period. <!-- from which estimator? -->The total biomass was further converted to an estimate of the total number of Chinook Salmon consumed by dividing by a weight of 10.9 g for the spring and 10.3 g for the fall, an average from juvenile Chinook Salmon captured during those date ranges at seven rotary screw traps located throughout the Upper Salmon upstream of Deadwater Slough.

<!-- Sawtooth Hatchery, East Fork Salmon, Pahsimeroi, Upper Lemhi, Hayden Creek, Lower Lemhi, North Fork Salmon -->
